---
title: "Checking spoilers Rmd"
author: "Evgeni Chasnovski"
date: 2017-09-25
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
library(dplyr)
```


`r blogdown::shortcode("spoiler", id = '\"sp1\"', title = '\"Spoiler-1\"')`

Some text.
```{r R chunk}
# This is comment
###' Second comment
#' Third comment
##Another comment
```

`r blogdown::shortcode("/spoiler")`

Some middle text.

`r blogdown::shortcode("spoiler", id = '\"sp2\"', title = '\"Spoiler-2\"')`
Another text.
`r blogdown::shortcode("/spoiler")`

`r blogdown::shortcode("spoiler", id = '\"sp3\"', title = '\"Spoiler-3\"')`

Text before code.

```{r Spoilered code}
# Be careful with comments inside spoilers
a <- 1:10

# They might be interpreted as markdown headers
## of any magnitude.

#After them theme highlighting doesn't work
#Simple workaround is to not leave whitespace between '#' and first word
#' Or just use roxygen comments.
b <- a * 2
```

`r blogdown::shortcode("/spoiler")`

Another text between spoilers.

`r blogdown::shortcode("spoiler", id = '\"sp4\"', title = '\"Spoiler-4\"')`

```{r Spoilered code without markdown text before it}
# However if there is no text before code

# Everything seems fine.
## Additional check
```

`r blogdown::shortcode("/spoiler")`