---
title: General gems of comeres
author: Evgeni Chasnovski
date: '2018-05-17'
publishDate: '2018-05-17'
slug: general-gems-of-comeres
categories: []
tags:
  - rstats
  - comperes
description: 'Examples of exported functions from comperes package that could be useful for general tasks.'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  collapse = TRUE,
  fig.width = 9
)
options(tibble.print_min = 5, tibble.print_max = 10)
```

# Prologue

I am very glad to announce that my new package [comperes](https://github.com/echasnovski/comperes) is [on CRAN](https://CRAN.R-project.org/package=comperes) now. It provides tools for managing competition results in a [tidy](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) manner as much as possible. For more information go to:

- Package [README](https://github.com/echasnovski/comperes#comperes-manage-competition-results).
- Package [vignettes](https://echasnovski.github.io/comperes/articles/).
- My [previous post](`r blogdown::shortcode(.name = "relref", '\"2018-05-09-harry-potter-and-competition-results-with-comperes.html\"', .type = "html")`) for usage examples based on built-in `hp_survey` data set (results of my [Harry Potter Books Survey](`r blogdown::shortcode(.name = "relref", '\"2018-04-09-struggle-with-harry-potter-data.html\"', .type = "html")`)).

Besides tools for competition results, `comperes` offers some functions that can be useful in more general tasks. This post presents examples of their most common usage.

# Overview

This post covers the following themes:

- __Compute vector levels__ with [levels2()](https://echasnovski.github.io/comperes/reference/levels2.html).
- __Manage item summaries__ with [summarise_item()](https://echasnovski.github.io/comperes/reference/item-summary.html) and [join_item_summary()](https://echasnovski.github.io/comperes/reference/item-summary-join.html).
- __Convert pairwise data__ with [long_to_mat() and mat_to_long()](https://echasnovski.github.io/comperes/reference/convert-pair-value.html).

For examples we will use a shortened version of the everlasting `mtcars` data set. We will need the following setup:

```{r library, message = FALSE}
library(comperes)
library(rlang)
# For example analysis
library(dplyr)
library(tibble)

mtcars_tbl <- mtcars %>%
  rownames_to_column(var = "car") %>%
  select(car, cyl, vs, carb) %>%
  as_tibble()
```

# Compute vector levels

We will start with the most simple function. During `comperes` development, idea about the it really helped me reason more clearly about package functional API. I am talking about [levels2()](https://echasnovski.github.io/comperes/reference/levels2.html) which computes "levels" of any non-list vector.

It has the following logic: if `x` has `levels` attribute then return `levels(x)`; otherwise return character representation of vector's sorted unique values. Notes about design and implementation of this function:

- I hesitated _a lot_ about whether it should return character or same type as input vector in case `x` has no `levels`. In many practical cases there is a need in latter behavior. However, in the end I decided that type stable output (`levels(x)` always returns character vector or `NULL`) is better.
- Conversion to character is done __after__ sorting, which is really important when dealing with numeric vectors.

This function is helpful when one needs to produce unique values in standardized manner (for example, during pairwise distance computation). Some examples:

```{r levels2}
levels2(mtcars_tbl$cyl)

# Importance of conversion to character after sorting
tricky_vec <- c(10, 1, 2, 12)
sort(as.character(tricky_vec))
levels2(tricky_vec)
```

# Manage item summaries

Arguably, the most common task in data analysis is computation of group summaries. This task is conveniently done by consecutive application of [dplyr](https://dplyr.tidyverse.org/)'s `group_by()`, `summarise()` and `ungroup()` (to return regular data frame and not grouped one). `comperes` offers a wrapper [summarise_item()](https://echasnovski.github.io/comperes/reference/item-summary.html) for this task (which always returns [tibble](https://tibble.tidyverse.org/) instead of a data frame) with additional feature of modifying column names by adding prefix (which will be handy soon):

```{r summarise_item}
cyl_vs_summary <- mtcars_tbl %>%
  summarise_item(
    item = c("cyl", "vs"),
    n = n(), mean_carb = mean(carb),
    .prefix = "cyl_vs__"
  )
cyl_vs_summary
```

Sometimes, there is also a need to compare actual values with their summaries across different grouping. For example, __determine whether car's number of carburetors (`carb`) is bigger than average value per different groupings__: by number of cylinders `cyl` and V/S `vs`.

To simplify this task, `comperes` offers a [join_item_summary()](https://echasnovski.github.io/comperes/reference/item-summary-join.html) function for that: it computes item summary with `summarise_item()` and joins it (with `dplyr::left_join()`) to input data frame:

```{r join_item_summary}
# Save (with rlang magic) expression for reused summary
carb_summary <- list(mean_carb = expr(mean(carb)))

# Create new columns with joined grouped summaries
mtcats_gear_summary <- mtcars_tbl %>%
  join_item_summary("cyl", !!! carb_summary, .prefix = "cyl__") %>%
  join_item_summary("vs",  !!! carb_summary, .prefix = "vs__")

print(mtcats_gear_summary, width = Inf)

# Compute comparisons
mtcats_gear_summary %>%
  mutate_at(vars(ends_with("mean_carb")), funs(carb > .)) %>%
  select(car, ends_with("mean_carb")) %>%
  rename_at(vars(-car), funs(gsub("__mean_carb$", "", .)))
```

Adding different prefixes helps navigating through columns with different summaries.

# Convert pariwise data

One of the main features of `comperes` is the ability to compute [Head-to-Head](https://echasnovski.github.io/comperes/reference/h2h_long.html#head-to-head-value) values of players in competition. There are functions `h2h_long()` and `h2h_mat()` which produce output in "long" (tibble with row describing one ordered pair) and "matrix" (matrix with cell value describing pair in corresponding row and column) formats respectively.

These formats of pairwise data is quite common: "long" is better for tidy computing and "matrix" is better for result presentation. Also converting distance matrix to data frame with pair data is a theme of several Stack Overflow questions (for example, [this one](https://stackoverflow.com/questions/27892100/distance-matrix-to-pairwise-distance-list-in-r) and [that one](https://stackoverflow.com/questions/21180464/distance-matrix-to-data-frame-pairs-in-r)).

Package `comperes` has functions `as_h2h_long()` and `as_h2h_mat()` for converting between those formats. They are powered by a "general usage" functions [long_to_mat() and mat_to_long()](https://echasnovski.github.io/comperes/reference/convert-pair-value.html). Here is an example of how they can be used to convert between different formats of pairwise distances:

```{r convert-pairwise-data}
# Compute matrix of pairwise distances based on all numeric columns
dist_mat <- mtcars_tbl %>%
  select_if(is.numeric) %>%
  dist() %>%
  as.matrix()
dist_mat[1:4, 1:4]

# Convert to data frame (tibble in this case)
dist_tbl <- dist_mat %>%
  mat_to_long(row_key = "id_1", col_key = "id_2", value = "dist")
dist_tbl

# Convert tibble back to matrix
dist_mat_new <- dist_tbl %>%
  # To make natural row and column sortings
  mutate_at(vars("id_1", "id_2"), as.numeric) %>%
  long_to_mat(row_key = "id_1", col_key = "id_2", value = "dist")
identical(dist_mat, dist_mat_new)
```

# Conclusion

- Package `comperes` provides not only tools for managing competition results but also functions with general purpose:
    - Compute vector levels with `levels2()`. Usually used to produce unique values in standardized manner.
    - Manage item summaries with `summarise_item()` and `join_item_summary()`. May be used to concisely compute comparisons of values with summaries from different groupings.
    - Convert pairwise data with `long_to_mat()` and `mat_to_long()`. Very helpful in converting pairwise distances between "long" and "matrix" formats.

`r blogdown::shortcode("spoiler", id = '\"sessionInfo\"', title = '\"sessionInfo()\"')`

```{r sessionInfo, eval = TRUE}
sessionInfo()
```

`r blogdown::shortcode("/spoiler")`