---
title: Transformers, glue!
author: Evgeni Chasnovski
date: '2018-08-21'
publishDate: '2018-08-21'
slug: transformers-glue
categories: []
tags:
  - rstats
  - glue
description: 'General description of transformers functionality in {glue} with some potentially useful examples.'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  collapse = TRUE,
  fig.width = 9, fig.height = 8
)
options(tibble.print_min = 5, tibble.print_max = 16)
```

# Prologue

Package {[glue](https://glue.tidyverse.org/index.html)} is designed as "small, fast, dependency free" tools to "glue strings to data in R". To put simply, it is a concise and flexible alternative for `paste()` with some additional features:

```{r glue-examples}
library(glue)

x <- 10
glue("I have {x} apples.")
paste("I have", x, "apples.")
```

Recently fate lead me to try using {glue} in a package. I was very pleased to how it makes code more readable, which I believe is a very important during package development. However, I stumbled upon this pretty unexpected behavior:

```{r glue-example-NULL}
y <- NULL
str(glue("I have {x} apples and {y} oranges."))
paste("I have", x, "apples and", y, "oranges.")
```

If one of the expressions is evaluated into `NULL` then the output becomes empty string. This was unintuitive result and for a while I thought about stop using {glue} as `NULL` is expected to be a valid input. However, if [Jim Hester](https://github.com/jimhester) is package author, you should expect some sort of designed solution to any problem. This time wasn't an exception with [transformers](https://glue.tidyverse.org/articles/transformers.html) functionality.

Basically, transformer is a function that changes the output of R expressions the way you want. As I wanted to make `NULL` visible, this is a perfect way to do it.

# Overview

This post describes an easy way to create {glue} wrappers with custom transformers and lists some examples that can be helpful in common tasks:

- **Transformers** uses a little bit of functional programming magic to create a potentially useful transformers.

Setup is very simple this time:

```{r library}
# {glue} was loaded in examples
# For functional programming magic
library(purrr)
# For string manipulation in one of the examples
library(stringr)
```

# Transformers

The task of creating wrapper for `glue()` essentially consists from two parts:

- Evaluate properly a supplied R expression.
- Modify it to show intended behavior.

The `transforming_glue()` wrapper does exactly this:

```{r transforming_glue}
transforming_glue <- function(transformer) {
  function(..., .sep = "", .envir = parent.frame(), .open = "{", .close = "}",
           .na = "NA") {
    glue(
      ..., .sep = .sep, .envir = .envir, .open = .open, .close = .close,
      .na = "NA",
      .transformer = compose(transformer, identity_transformer)
    )
  }
}
```

Breakdown of this code:

- Input is a `transformer` - function that takes an already evaluated R object and modifies it the way you want.
- Output is a function that is a wrapper for `glue()`. Its transformer is a function composition that first evaluates R expression with `identity_transformer` (function from {glue}) and then applies supplied `transformer`. Composition here is done with [compose()](https://purrr.tidyverse.org/reference/compose.html) - an element of functional programming magic from {[purrr](https://purrr.tidyverse.org/)}.

## Show NULL

Back to initial problem. We want `NULL` to be a valid R value for a `glue()`:

```{r show-null}
show_null <- function(x, val = "NULL") {
  if (is.null(x)) {
    val
  } else {
    x
  }
}

glue_null <- transforming_glue(show_null)

# Example from Prologue
glue_null("I have {x} apples and {y} oranges.")
```

## Fixed width output

With {[stringr](https://stringr.tidyverse.org/)} package you can force an output to be fixed width:

```{r fixed-width}
str_width <- function(x, width) {
  if (str_length(x) > width) {
    str_trunc(x, width, side = "right")
  } else {
    str_pad(x, width, side = "right")
  }
}

glue_width <- transforming_glue(partial(str_width, width = 10))

short_oh <- "Ooh!"
long_oh <- "Oooooooooooh!"
glue_width("This puzzles ({short_oh}) and surprises ({long_oh}) me.")
```

**Note** usage of [partial()](https://purrr.tidyverse.org/reference/partial.html) here: it takes function along with its arguments values and modifies it by "pre-filling" those arguments.

## Enclose output

In some situation you might want to explicitly show which strings represent R objects in the output. You can do that by enclosing the output in some sort of braces:

```{r enclose-output}
enclose <- function(x, start = "<", end = ">") {
  paste0(start, x, end)
}

glue_enclose <- transforming_glue(enclose)

glue_enclose("What if I had {x} oranges?")
```

## Bizarro encryption

One possibly useful pattern is to encrypt the used data to prevent it from seeing by untrustworthy eyes. Here we will use simplified `bizarro()` example from [this insightful UseR 2018 talk](https://www.youtube.com/watch?v=7oyiPBjLAWY) by the amazing [Jennifer (Jenny) Bryan](https://github.com/jennybc). Here `glue_bizarro()` "reverts" R objects based on their type.

```{r bizarro}
str_reverse <- function(x) {
  vapply(
    strsplit(x, ""),
    FUN = function(z) paste(rev(z), collapse = ""),
    FUN.VALUE = ""
  )
}

bizarro <- function(x) {
  cls <- class(x)[[1]]
  
  switch(
    cls,
    logical = !x,
    integer = -x,
    numeric = -x,
    character = str_reverse(x),
    x
  )
}

glue_bizarro <- transforming_glue(bizarro)

new_fruit <- "pomegranate"
glue_bizarro(
  "Then I might have {x + 10} apples. Is that {TRUE}?
   Maybe I want {new_fruit}?"
)
```

## Ultimate example

Using already familiar functional programming technique, we can create an ultimate `glue()` wrapper as a combination, or rather `compose()`-ition, of all previous examples. The most important part is supply the in correct order:

```{r ultimate}
glue_ultimate <- transforming_glue(
  compose(
    enclose,
    partial(str_width, width = 10),
    # To ensure that input of `str_width()` is character
    as.character,
    show_null,
    bizarro
  )
)

glue_ultimate(
  "I have {x} apples and {y} oranges.
   This puzzles ({short_oh}) and surprises ({long_oh}) me.
   What if I had {x} oranges?
   Then I might have {x + 10} apples. Is that {TRUE}?
   Maybe I want {new_fruit}?"
)
```

# Conclusions

- Package {glue} is a very useful and flexible way of creating strings based on evaluation of R expressions.
- Its "transformer" functionality is an interesting way to manipulate string output by supplying custom modification function.
- Functional programming with {purrr} can be very helpful in creating concise and extensible code.

`r blogdown::shortcode("spoiler_details", summary = '\"sessionInfo()\"')`

```{r sessionInfo, eval = TRUE}
sessionInfo()
```

`r blogdown::shortcode("/spoiler_details")`